---
title: "Global plastic waste"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

Plastic pollution is a major and growing problem, negatively affecting oceans and wildlife health. [Our World in Data](https://ourworldindata.org/plastic-pollution) has a lot of great data at various levels including globally, per country, and over time. For this lab we focus on data from 2010.

Additionally, National Geographic ran a data visualization communication contest on plastic waste as seen [here](https://www.nationalgeographic.com/environment/article/plastic-innovation-challenge-prize-clean-ocean).

# Learning goals

-   Visualize numerical and categorical data and interpret your visualisations.
-   Recreate visualisations.

## Packages

We'll use the **tidyverse** package for this analysis. Run the following code in the Console to load this package.

```{r load-packages, message=FALSE, eval=TRUE}
library(tidyverse)
library(here)
library(patchwork)
```

## Data

The dataset for this assignment can be found as a csv file. You can read it in using the following.

```{r load-data, message=FALSE, eval=TRUE}
plastic_waste <- read_csv(here("1_plastic-waste","data", "plastic-waste.csv"))
```

The variable descriptions are as follows:

-   `code`: 3 Letter country code
-   `entity`: Country name
-   `continent`: Continent name
-   `year`: Year
-   `gdp_per_cap`: GDP per capita constant 2011 international \$, rate
-   `plastic_waste_per_cap`: Amount of plastic waste per capita in kg/day
-   `mismanaged_plastic_waste_per_cap`: Amount of mismanaged plastic waste per capita in kg/day
-   `mismanaged_plastic_waste`: Tonnes of mismanaged plastic waste
-   `coastal_pop`: Number of individuals living on/near coast
-   `total_pop`: Total population according to Gapminder

# Warm up within your pair

You have 5 minutes to complete this section.

**Warm up 1:** Verify that the dataset has loaded into the Environment. How many observations are in the dataset? Clicking on the dataset in the Environment will allow you to inspect it more carefully. Alternatively, you can type `View(plastic_waste)` into the Console to do this.
- 240 observations

**Warm up 2:** Have a quick look at the data and notice that there are cells taking the value `NA` -- what does this mean?
- there no data for those cells
# Exercises

Let's start by taking a look at the distribution of plastic waste per capita in 2010.

```{r plastic_waste_per_cap-hist, eval=TRUE}
x = plastic_waste_per_cap)) +
  geom_histogram(binwidth = 0.2)
```

One country stands out as an unusual observation at the top of the distribution. One way of identifying this country is to filter the data for countries where plastic waste per capita is greater than 3.5 kg/person.

```{r plastic_waste_per_cap-max, eval=TRUE}
plastic_waste %>%
  filter(plastic_waste_per_cap > 3.5)
```

Did you expect this result? You might consider doing some research on Trinidad and Tobago to see why plastic waste per capita is so high there, or whether this is a data error.
- surpised

1.  Plot, using histograms, the distribution of plastic waste per capita faceted by continent. What can you say about how the continents compare to each other in terms of their plastic waste per capita?

```{marginfigure}
plastic_waste %>% 
  ggplot(aes(x=plastic_waste_per_cap))+
  geom_histogram()+
  facet_wrap(~continent)

**NOTE:** From this point onwards the plots and the output of the code are not displayed in the lab instructions, but you can and should write the code and view the results yourself.

```

Another way of visualizing numerical data is using density plots.

```{r plastic_waste_per_cap-dens}
ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap)) +
  geom_density()
```

And compare distributions across continents by coloring density curves by continent.

```{r plastic_waste_per_cap-dens-color}
ggplot(data = plastic_waste, 
       mapping = aes(x = plastic_waste_per_cap, 
                     color = continent)) +
  geom_density()
```

The resulting plot may be a little difficult to read, so let's also fill the curves in with colors as well.

```{r plastic_waste_per_cap-dens-color-fill}
ggplot(data = plastic_waste, 
       mapping = aes(x = plastic_waste_per_cap, 
                     color = continent, 
                     fill = continent)) +
  geom_density()
```

The overlapping colors make it difficult to tell what's happening with the distributions in continents plotted first, and hence coverred by continents plotted over them. We can change the transparency level of the fill color to help with this. The `alpha` argument takes values between 0 and 1: 0 is completely transparent and 1 is completely opaque. There is no way to tell what value will work best, so you just need to try a few.

```{r plastic_waste_per_cap-dens-color-fill-alpha}
ggplot(data = plastic_waste, 
       mapping = aes(x = plastic_waste_per_cap, 
                     color = continent, 
                     fill = continent)) +
  geom_density(alpha = 0.3)
```

This still doesn't look great...

2.  Recreate the density plots above using a different (lower) alpha level that works better for displaying the density curves for all continents.

3.  Describe why we defined the `color` and `fill` of the curves by mapping aesthetics of the plot but we defined the `alpha` level as a characteristic of the plotting geom.
- becuase alpha is an argument within the geom function not the aes function 

And yet another way to visualize this relationship is using side-by-side box plots.

```{r plastic_waste_per_cap-box}
ggplot(data = plastic_waste, 
       mapping = aes(x = continent, 
                     y = plastic_waste_per_cap)) +
  geom_boxplot()

{r plastic_waste_per_cap-box}
ggplot(data = plastic_waste, 
       mapping = aes(x = continent, 
                     y = plastic_waste_per_cap)) +
  geom_violin()
```

4.  Convert your side-by-side box plots from the previous task to [violin plots](http://ggplot2.tidyverse.org/reference/geom_violin.html). What do the violin plots reveal that box plots do not? What features are apparent in the box plots but not in the violin plots?
- in violin plot you don't get outliers, median, IQR 

5.  Visualize the relationship between plastic waste per capita and mismanaged plastic waste per capita using a scatterplot. Describe the relationship.
```{r plastic_waste_per_cap-box}
plastic_waste %>% 
  ggplot(aes(x=mismanaged_plastic_waste_per_cap, y= plastic_waste_per_cap, colour = continent))+
  geom_point()
# slightly positive correlation between mismanaged plastic waste per cap and plastic waste per capita
```
6.  Color the points in the scatterplot by continent. Does there seem to be any clear distinctions between continents with respect to how plastic waste per capita and mismanaged plastic waste per capita are associated?
- Yes, for example, Europe has high plastic waste per capita but relatively low mismanaged plastic waster per capita and there little correlation within North America between the two variables

7.  Visualize the relationship between plastic waste per capita and total population as well as plastic waste per capita and coastal population. You will need to make two separate plots. Do either of these pairs of variables appear to be more strongly linearly associated?
```{r plastic_waste_per_cap-box}
x <- plastic_waste %>% 
  ggplot(aes(x = plastic_waste_per_cap, y = total_pop))+
  geom_point() 

y <- plastic_waste %>% 
  ggplot(aes(x = plastic_waste_per_cap, y = coastal_pop))+
  geom_point()

x + y
# patchwork package to put the two plots side by side on one screen in the viewer

# maybe more of a linear association when comparing coastal population, but this is difficult to determine due to outliers

```
